Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-08-28 paper/manuscript.tex
Updates from Overleaf

\affil[1]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[2]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{National Institutes of Health, Bethesda, MD, USA}
\affil[5]{IBM Alamden Research Center, Almaden, CA, USA}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[11]{Computational Chemistry, Janssen Research & Development, Turnhoutseweg 30, Beerse B-2340,Belgium}

@@

%%%%%%%%%%%%
% Step 0   %
%%%%%%%%%%%%
\section{What can be expected from alchemical simulations? -- Step 0}
\label{sec:step0}
\todo[inline, color={green!20}]{@everyone please review this section}
Step 0 is about initially counting the cost of your
project: Can you even hope to tackle the problem you are thinking
about attempting given available resources and, if successful, will it
be worth it?
\subsection*{How accurate are current alchemical free energy calculations?}
Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative~\cite{lots-of-citations}. 
\todo[inline,color={green!20}]{JDC: We should probably provide a short list of key benchmark references here, broken down by force field, and keep this up to date. GT: would a table, at the end of the document be an easier way to do this, only requiring some brief comments in the text? ASJSM: I like the table idea.}
However, the domain of applicability is a significant concern~\cite{Sherborne2016, Cournia2017}, especially for relative calculations, which typically require a bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\subsection*{Is the expected accuracy of the computation good enough?}
Additionally, the requisite level of accuracy is important -- if the
goal is to guide lead optimization when many compounds will be
synthesized, free energy calculations can be appealing even with
accuracies in the 1-2 kcal/mol range~\cite{brown2010freeenergy,mobley2012perspective}, but if the number of compounds to be
synthesized is very small, this accuracy may not be enough to provide
much value.

@@

  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right)

@@

probability

@@

  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0)

@@

of free energy calculations is $\sigma = 1$kcal/mol, $P(\Delta\Delta

@@

\item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.  Specifically, if the $\lambda$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e. between $i$ and $i+1$ and between $i$ and $i-1$, they will \testit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$. Thus MBAR doesn't actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal.benchmark} 

@@

\subsubsection{Are my simulations any good?}
\todo[inline, color={green!20}]{ASJSM: write and tidy}

\todo[inline, color={yellow!20}]{@MKuhn: write section on overlap with example plots}
\begin{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Other considerations for many transformations
\end{itemize}

@@

\begin{figure}
    \includegraphics[width=0.95\linewidth]{fig6/free_energy_trajectories.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{Samar: write section}
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schrodinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format. 
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM sctipts for solvation and ligand-binding free energy calculations. 
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implemente. The LiveCOMS GROMACS tutorial has an example free energycalculation~\cite{LiveCOMS-gromacs}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

@@

>>>>

\affil[1]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[2]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{National Institutes of Health, Bethesda, MD, USA}
\affil[5]{IBM Alamden Research Center, Almaden, CA, USA}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[11]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
\affil[12]{Cresset, Cambridgeshire, UK}

@@

%%%%%%%%%%%%
% Step 0   %
%%%%%%%%%%%%
\section{What can be expected from alchemical simulations? -- Step 0}
\label{sec:step0}
\todo[inline, color={green!20}]{@everyone please review this section}
Step 0 is about initially counting the cost of your
project: Can you even hope to tackle the problem you are thinking
about attempting given available resources and, if successful, will it
be worth it?
\subsection*{How accurate are current alchemical free energy calculations?}
Current alchemical free energy calculations seem to achieve, in favorable cases, RMS errors around 1-2 kcal/mol depending on force field, system, and a variety of other factors such as simulation time, sampling method, and whether the calculations employed are absolute or relative. An overview of benchmark datasets and case studies can be found in section \ref{sec:benchmark} at the end of this document.
\todo[inline,color={green!20}]{JDC: We should probably provide a short list of key benchmark references here, broken down by force field, and keep this up to date. GT: would a table, at the end of the document be an easier way to do this, only requiring some brief comments in the text? ASJSM: I like the table idea. MK: Feel free to remove this todo item if you are happy with the table provided. I refrained from just listing lots of case studies, as I think we should highlight the recommended datasets which should be used for benchmarking. Nevertheless, I have cited an review article containing such a summary as well as a recent publication comparing FEP+ and Gromacs in the brief text below the table.}
However, the domain of applicability is a significant concern~\cite{Sherborne2016, Cournia2017}, especially for relative calculations, which typically require a bound structure of a closely
related ligand as a starting point. Additional factors such as slow protein or ligand rearrangements, uncertainties in ligand binding mode, or charged ligands can make these calculations far less reliable and more of a research effort.

@@

\subsection*{Is the expected accuracy of the computation good enough?}
Additionally, the requisite level of accuracy is important -- if the
goal is to guide lead optimization when many compounds will be
synthesized, free energy calculations can be appealing even with
accuracies in the 1-2 kcal/mol range~\cite{mobley2012perspective}, but if the number of compounds to be
synthesized is very small, this accuracy may not be enough to provide
much value.

@@

  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),

@@

probability:

@@

  P(\Delta\Delta G<0|\Delta\Delta G^\dagger<0).

@@

of free energy calculations is $\sigma = 1$ kcal/mol, $P(\Delta\Delta

@@

\item BAR is essentially just as good as MBAR for highly optimized $\lambda$ intermediates.  Specifically, if the $\lambda$s are chosen such that intermediate states have moderate overlap with their neighbors (i.e. between $i$ and $i+1$ and between $i$ and $i-1$, they will \textit{not} have significant overlap with their next nearest neighbors $i+2$ and $i-2$. Thus MBAR doesn't actually get significant information from these energy differences, so one might as well not even calculate them, and just perform BAR between nearest neighbors.~\cite{paliwal.benchmark} 

@@

\subsubsection{Are my simulations any good?}
\todo[inline, color={green!20}]{ASJSM: write and tidy}
One way of assessing reliability of the calculations may be by checking the phase space overlap between neighboring $\lambda$-windows \cite{wu2005phasespace,wu2005phasespacea}. For this purpose, a so-called overlap matrix can be used. This is a $K\times K$ matrix, where $K$ is the number of simulated states, i.e. $\lambda$s. Sufficient overlap is important for reweighting estimators such as BAR or MBAR, but cannot help assess reliability of estimates, when using TI. 
These matrices are graphical representations of the phase space overlap, i.e. the average probability that a sample generated at state $j$ can be observed at state $i$. As this probability is computed considering the samples from all states (and not just the adjacent states), the values in each row and column add up to 1. Details on the calculation and properties of these matrices can be found elsewhere~\cite{klimovich2015guidelines}.
In an overlap matrix $\mathbf{O}$, the off-diagonal values (${O}_{i,j,i\ne j}$) are negatively correlated with the variance of the free energy difference. Accordingly, the uncertainty of the free energy difference between the states i and j will be smaller when ${O}_{i,j,i\ne j}$ is larger (and thus the values in the main diagonal (${O}_{i,j,i=j}$) are smaller). In order to obtain a reliable estimate of the free energy all neighbouring states must be connected, i.e. there must be sufficient overlap between the samples of these states (general description: ${O}_{i,j,i\ne j}\ge$ threshold).
However, due to the mathematical derivation it is difficult to explicitly describe the relation of the overlap matrix and the variance by formulae. Consequently, the threshold has to be derived empirically. It has been proposed that the values of the first off-diagonals (i.e. the diagonals above and below the main diagonal) should at least be 0.03 to obtain a reliable free energy estimate~\cite{klimovich2015guidelines}. Smaller values should be considered as a warning sign (see Figure~\ref{fig:overlap}), as the variance tends to be underestimated in case of poor overlap.
\todo[inline, color={green!20}]{ASJSM: Please upload the updated figure for the overlap matrices and modify the path below, if necessary}
\begin{figure}
\includegraphics[width=0.95\columnwidth]{figures/overlap/Overlap_plot.pdf}
\caption{\label{fig:overlap}Overlap matrices showing (A) good, (B) mediocre, and (C) poor overlap.}
\end{figure}
Figure~\ref{fig:overlap} shows examples of good, mediocre and poor overlap in ~\ref{fig:overlap} A, B, and C respectively. For (A), the probability to find a sample from state $i$ in its neighbouring state $j$ is about 0.2 for all states adjacent to the main diagonal, and hence the overall connectivity is good. In the case of (B), the overlap is strongly diminishing in the lower right corner, raising concerns regarding the reliability of the free energy estimate obtained. For (C), the state at $\lambda$ index = 6 is connected to neither of its neighbouring states. While this does not necessarily imply that the result for this perturbation is wrong, the energy estimate must at least be considered as highly unreliable.
In order to overcome the issue of poor overlap, additional sampling should be performed by introducing additional states, i.e. $\lambda$ windows.
Interestingly, as the variance is inversely correlated with the number of states\cite{klimovich2015guidelines}, it can in principle be reduced below any arbitrary threshold. However, decreasing the variance to a value close to 0 is not feasible, as this approach would strongly increase the calculation time. Instead, the uncertainty may be decreased most effectively by adjusting the $\lambda$ spacing to perform additional sampling in the areas of poor overlap, rather than using arbitrarily spaced windows. Different approaches are described in section \ref{part_of_section_six}) and more details can be found in the literature~\cite{dakka2018concurrent, hahn2019alchemical}.
On a final note, it has been argued that overlap matrices are only a qualitative estimator for the spacing of windows. Another estimator named Time Derivative of overall Variance (TDV)~\cite{sun2017barbased} has been suggested for this task, which is an explicit function of the overall variance. However, it could not be proven that better suggestions for the $\lambda$ spacing can be obtained by this approach compared to overlap matrices. Hence, it should rather be seen as a complementary approach. 
In a nutshell, it must be ensured that the values of an overlap matrix in the first diagonals above and below the main diagonal are at least 0.03 in order to yield reliable results.

\begin{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Other considerations for many transformations
\end{itemize}

@@

\todo[inline, color={green!20}]{ASJSM: Please upload the correct figure here and modify the path below}
\begin{figure}
    \includegraphics[width=0.95\linewidth]{fig6/free_energy_trajectories.pdf}
    \caption{Average binding free energy of 5 replicate Hamiltonian replica exchange calculations as a function of total simulation time (i.e. the sum of the simulation time of all replicas) for the two host-guest systems CB8-G3 and OA-G3. Shaded areas represent 95\% confidence intervals around the mean computed from the 5 replicates data. The horizontal dashed lines show the final binding free energy prediction of the two calculations after a total of ~5230 ns for OA-G3 and ~6650 ns for CB8-G3. Longer correlation times in CB8-G3 cause the calculation to converge more slowly. The original data used to generate the plot can be found at \url{https://github.com/MobleyLab/SAMPL6/blob/master/host_guest/Analysis/SAMPLing/Data/reference_free_energies.csv}.
}
    \label{fig:freeenergytrajectories}
\end{figure}

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{Samar: write section}
\label{sec:software}
There are several softwares availble for carrying out alchemical free energy calculations. In this section we discuss some of the popular commercial and free tools along with their features. 
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+ is a tool offered by Schrodinger Inc. under a commercial license. It has an intuitive GUI which makes it easier for non-experts to run alchemical free energy calculations and analyze the results. It runs DESMOND MD package under the hood and hence parallelizes very well on the GPUs. 
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM has a variety of tools developed over the years. PERT module can be used to define initial and final states and define the intermediate lambda points. FREN and BAR modules can be used to analyze the data after the MD run. Lambda-dynamics based free energy calulation can be carried out using the BLOCK module.  
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER, including its new pmemd.cuda version support free energy calculation. 
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item PLUMED is an open source tool which enables the usage of a variety of MD engines. It is designed as a plugin for MD packages such that it analyzes the trajectory on the fly. It also offers a VMD based plugin for the computation of collective variables.   	
	\item SIRE
	\item YANK is a tool developed by John Chodera and group on the top of OpenMM MD package. It allows the users to write their inputs in easy-to-use YAML format.
	\item GROMACS is a molecular simulation package with a significant number of free energy methods implementations. The LiveCOMS GROMACS tutorial has an example free energy calculation~\cite{LiveCOMS-gromacs}.
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\item CHARMM-GUI is a web based tool for setting up a variety of MD simulations. It can be used to generate CHARMM sctipts for solvation and ligand-binding free energy calculations.
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

@@

\clearpage
\section{Benchmark Datasets}
\label{sec:benchmark}
\vspace{5mm}
\begin{tabular}{cccc}
\caption{Overview of benchmark datasets}
\textbf{Publication} & \textbf{Targets} & \textbf{Ligands} & \textbf{Force Field} \\
\hline
\multicolumn{4}{|c|}{D3R Grand Challenges \cite{D3R}} \\
\hline
GC3 \cite{gaieb2019} & 6 & 266 & various \\
GC2 \cite{gaieb2018} & 1 & 102 & various \\
GC2015 \cite{gathiaka2016} & 2 & 215 & various \\
\hline
\multicolumn{4}{|c|}{SAMPL Challenges \cite{SAMPL}} \\
\hline
SAMPL6 \cite{rizzi2018} & 3 & 21 & various \\
SAMPL5 \cite{yin2017} & 3 & 22 & various \\
SAMPL4 \cite{muddana2014} & 2 & 23 & various \\
\hline
\multicolumn{4}{|c|}{Schrödinger Datasets} \\
\hline
FEP+ Dataset \cite{wang2015} & 8 & 199 & OPLS2.1 \\
FEP+ Dataset \cite{harder2016} & 8 & 199 & OPLS3 \\
FEP+ Dataset \cite{roos2019} & 8 & 199 & OPLS3e \\
FEP+ Dataset \cite{song2019} & 8 & 199 & AMBER18 \\
Fragment Optimization \cite{steinbrecher2015} & 8 & 96 & OPLS2.1 \\
Scaffold Hopping \cite{wang2017} & 6 & 21 & OPLS3 \\
Macrocycles \cite{yu2017} & 7 & 33 & OPLS3 \\
\hline
\multicolumn{4}{|c|}{Further Suggested Datasets} \\
\hline
Cucurbit[7]uril (CB7) \cite{mobley2017} & 1 & 15 & NA \\
Deep cavity cavitand \cite{mobley2017} & 2 & 19 & NA \\
T4 Lysozyme \cite{mobley2017} & 2 & 20 & NA \\
\hline
\label{tab:benchmarks}
\end{tabular}

\todo[inline, color={green!20}]{ASJSM: Please check whether you can format this in a more appealing way.}
Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al. \cite{williams2017}. For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf. the recently published study by Pérez-Benito et al. \cite{perez2019}.
An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson \cite{mobley2017} or on alchemistry.org \cite{alchemistry}. These include cyclodextrins, the Cytochrome c peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets \cite{paliwal2011}.
